---
title: "Deriving Insight From Lahman's Baseball Database"
author: "Michael Calabro"
date: "7/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>
Welcome to my journey_to_insight.Rmd file! In this markdown, I will be going step by step through my process of using the Lahman Baseball Database to seek answers to my question regarding Strikeouts in Major League Baseball.  
<br>  
<center>
**What is the Lahman Baseball Database? And what is your question about MLB strikeouts?**
</center>
<br>
The Lahman Baseball Database contains MLB statistics and data from 1871-2019...it contains nearly 250 years worth of MLB data to explore!  Documentation for the database and all of its tables and columns can be found using [this link](http://www.seanlahman.com/files/database/readme2019.txt).  
I downloaded the SQLite database into my RStudio Project, and can access all of its information with the following commands.

```{r, warning=FALSE, message=FALSE}
# RSQLite library allows me to use SQLite
# DBI library allows me to connect to the database
# tidyverse contains many libraries, most notably ggplot2 for visualizations
# kableExtra allows me to make my tables more presentable
library(RSQLite)
library(DBI)
library(tidyverse)
library(kableExtra)

# Now I need to establish my connection to the Lahman database
con <- dbConnect(SQLite(), 
                 dbname = "lahmans_baseball_db.sqlite")
```
<br>  
This "con" keyword now allows me to connect to the database. To view the list of tables in the database, I simply run the code below.  

```{r}
dbListTables(con)
```
<br>  
To view data from a specific table, I use the dbGetQuery function. The "batting" table will be key to my analysis, let's check it out.

```{r}
# First I write out the query and assign it to "query"
query <-  "
          SELECT *
          FROM batting
          LIMIT 5
          "

# Then I send it through the DBGetQuery function, along with the "con" connection, to run the query
dbGetQuery(con, query)

# And for tables that aren't so wide, the kable funtion with kable_styling makes the table very presentable
dbGetQuery(con, query) %>%
  kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "bordered")

# I like the kable and think I'll use it a lot, so I am going to make a function kable_query for efficiency
kable_query <- function(con, query){
  return(
    dbGetQuery(con, query) %>%
      kable() %>%
      kable_styling(full_width = FALSE, bootstrap_options = "bordered")
  )
}
```
<br>  
As we can see, every row in the batter table consists of a player, a year, and all of the player's batting statistics in that year.  
<br>
<center>
**So What is your question about Strikeouts in the MLB?**
</center>
<br>  
A few years back (in 2018 I believe) I read an article which mentioned the fact that MLB players are striking out more and more every year, always breaking the single season record for total strikeouts. A more updated article, linked [here](https://www.espn.com/mlb/story/_/id/27694758/mlb-sets-strikeout-record-12th-straight-season), notes that this increase in strikeouts has been happening every year since 2008! Let's see if our Lahman data shows the same trend.

```{r}
query <-  "
          SELECT 
            yearID,
            SUM(SO) AS total_strikeouts
          FROM batting
          WHERE yearID > 2006
          GROUP BY yearID
          "

kable_query(con, query)

# Now I can use ggplot, with geom_point to make a scatterplot, geom_line to make a line graph, or both!
# Also, using %>% allows me to "pipe" my data into certain functions, like ggplot
dbGetQuery(con, query) %>%
  ggplot(aes(x = yearID, y = total_strikeouts)) +
  geom_point() +
  geom_line() +
  ggtitle("Yearly Total MLB Strikeouts From 2007-2019")
```

So what is causing this increase in strikeouts every year?? That is what I wish to investigate throughout the rest of this document/project.  
<br>
<center>
**HYPOTHOSES**
</center>
<br>

* Perhaps there are just more at bats every year, while strikeouts per at bat is remaining constant 

* Perhaps hitters are willing to strike out more often, in exchange for an increase in another statistic

* Perhaps pitching skill is improving faster than batting skill in the MLB  

<br>
<u>
<center>
**Strikeouts Per At Bat**
</center>
</u>
<br>  

This hypothesis is pretty easy to check. We just need to divide total strikeouts by total at bats.

```{r}
query <-    "
            SELECT 
              yearID,
              SUM(AB) AS total_at_bats,
              SUM(SO) AS total_strikeouts,
              CAST(SUM(SO) AS FLOAT) / SUM(AB) AS strikeouts_per_at_bat
            FROM batting
            WHERE yearID > 2006
            GROUP BY yearID
                  "

kable_query(con, query)

dbGetQuery(con, query) %>%
  ggplot(aes(x = yearID, y = strikeouts_per_at_bat)) +
  geom_line() +
  ggtitle("Yearly Strikeouts Per At Bat In The MLB From 2007-2019")
```

Now the obvious question is: Have any other hitting stats been rising or falling in lockstep with the strikeouts stat?

```{r}
query <-    "
            SELECT 
              yearID,
              CAST(SUM(SO) AS FLOAT) / SUM(AB) AS strikeouts_per_at_bat,
              CAST(SUM(H) AS FLOAT) / SUM(AB) AS hits_per_at_bat,
              CAST(SUM(HR) AS FLOAT) / SUM(AB) AS homeruns_per_at_bat,
              CAST(SUM('2B') AS FLOAT) / SUM(AB) AS doubles_per_at_bat,
              CAST(SUM('3B') AS FLOAT) / SUM(AB) AS triples_per_at_bat,
              CAST(SUM('3B') + SUM('2B') + SUM(HR) AS FLOAT) / SUM(AB) AS xbh_per_at_bat,
              CAST(SUM(BB) AS FLOAT) / SUM(AB) AS walks_per_at_bat
            FROM batting
            WHERE yearID > 2006
            GROUP BY yearID
            "

kable_query(con, query)
```

Homeruns and extra base hits (xbh) are also much higher in 2019 than they were in 2007, but the jump was not as consistent as the gradual rise of strikeouts.

```{r}
dbGetQuery(con, query) %>%
  ggplot(aes(x = yearID)) +
  geom_line(aes(y = xbh_per_at_bat, color = "xbh per at bat")) +
  geom_line(aes(y = homeruns_per_at_bat, color = "homeruns per at bat"))

dbGetQuery(con, query) %>%
  ggplot(aes(x = yearID)) +
  geom_line(aes(y = xbh_per_at_bat, color = "xbh per at bat")) +
  geom_line(aes(y = strikeouts_per_at_bat, color = "strikeouts per at bat"))
```

They don't appear to be too correlated. Let's check it out with a scatterplot.

```{r}
dbGetQuery(con, query) %>%
  ggplot(aes(x = xbh_per_at_bat, y = strikeouts_per_at_bat)) +
  geom_point() +
  # Stat_smooth, with geom=line and method=lm, adds a regression line of best fit to the graph
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")
```

There appears to be some correlation between strikeouts and extra base hits, but it is not a very strong one, and there is some heteroskedasticity.  
What if we go further back in time?

```{r}
query <-    "
            SELECT 
              yearID,
              CAST(SUM(SO) AS FLOAT) / SUM(AB) AS strikeouts_per_at_bat,
              CAST(SUM(H) AS FLOAT) / SUM(AB) AS hits_per_at_bat,
              CAST(SUM(HR) AS FLOAT) / SUM(AB) AS homeruns_per_at_bat,
              CAST(SUM('2B') AS FLOAT) / SUM(AB) AS doubles_per_at_bat,
              CAST(SUM('3B') AS FLOAT) / SUM(AB) AS triples_per_at_bat,
              CAST(SUM('3B') + SUM('2B') + SUM(HR) AS FLOAT) / SUM(AB) AS xbh_per_at_bat,
              CAST(SUM(BB) AS FLOAT) / SUM(AB) AS walks_per_at_bat
            FROM batting
            WHERE yearID > 1920
            GROUP BY yearID
            "

dbGetQuery(con, query) %>%
  ggplot(aes(x = yearID)) +
  geom_line(aes(y = xbh_per_at_bat, color = "xbh per at bat")) +
  geom_line(aes(y = strikeouts_per_at_bat, color = "strikeouts per at bat"))

dbGetQuery(con, query) %>%
  ggplot(aes(x = xbh_per_at_bat, y = strikeouts_per_at_bat)) +
  geom_point() +
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")

y <- dbGetQuery(con, query)$strikeouts_per_at_bat
x <- dbGetQuery(con, query)$xbh_per_at_bat
summary(lm(y ~ x))
```

There is certainly *some* correlation here - extra base hits and strikeouts have both been increasing in the MLB over time. But this doesn't confirm any specific hypothesis. What would we also expect to see as a trend?

* Intuitively, you would expect a player who strikes out more to have a lower salary, and someone who hits more extra base hits to have a higher salary. Perhaps extra base hits have a larger effect on salary than strikeouts, so players are incentivized to hit extra base hits at the expense of striking out more for the sake of making more money in contract negotiations.  

* I also want to look at what is more correlated with scoring runs - strikeouts or extra base hits. Perhaps extra base hits have a larger effect on scoring runs than strikeouts, so players are incentivized to hit extra base hits at the expense of striking out more for the sake of winning more games.

<br>
<u>
<center>
**Salary and Strikeouts**
</center>
</u>
<br> 

```{r}
query <- "
          SELECT *
          FROM salaries
          LIMIT 5
         "

kable_query(con, query)
```

```{r}
query <-  "
          SELECT 
            playerID || ' ' || yearID AS player_and_year,
            salary
          FROM salaries
          LIMIT 5
          "

kable_query(con, query)
```

```{r}
query <-  "
          SELECT
            playerID || ' ' || yearID AS player_and_year,
            playerID,
            yearID,
            teamID,
            AB,
            R,
            H,
            batting.'2B',
            batting.'3B',
            HR,
            RBI,
            BB,
            SO
          FROM batting
          LIMIT 5
          "

kable_query(con, query)
```

Okay we've hit a roadblock. I am trying to combine these two tables together, but when I add certain "Where" conditions, it gets super slow. So I am going to try using dplyr to help me "filter".  
That did not work either. 
Fixing it by limitting my rows and columns to 

```{r}
query <-  "
      WITH select_batting AS (
          SELECT 
            playerID || ' ' || yearID AS player_and_year,
            CAST(SO AS FLOAT) / AB AS strikeouts_per_at_bat,
            CAST(b.'3B' + b.'2B' + HR AS FLOAT) / AB AS xbh_per_at_bat,
            CAST(HR AS FLOAT) / AB AS hr_per_at_bat
          FROM batting AS b
          WHERE (AB > 300) AND (yearID IS 2016)
      ), select_salary AS (
          SELECT
            playerID || ' ' || yearID AS player_and_year,
            salary
          FROM salaries
          WHERE yearID IS 2016
      )
      SELECT 
        sb.*,
        ss.salary /CAST(1000000 AS FLOAT) AS salary_millions
      FROM select_batting AS sb
      INNER JOIN select_salary AS ss ON ss.player_and_year = sb.player_and_year
      WHERE salary_millions > 0
          "

head(dbGetQuery(con, query))

dbGetQuery(con, query) %>%
  ggplot(aes(x = strikeouts_per_at_bat, y = salary_millions)) +
  geom_point() +
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")

```

Obviously little to no correlation there. Interesting. To do this better I would want to know stats for a year before a contract is signed, but for now this is a fine approximation.

```{r}
dbGetQuery(con, query) %>%
  ggplot(aes(x = hr_per_at_bat, y = salary_millions)) +
  geom_point() +
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")
```

Not much there either.  

Now I'm going to see the correlation of strikeouts with runs scored, and the correlation of xbh with runs scroed for teams in certain years.

```{r}
query <-  "
          SELECT 
            yearID,
            teamID,
            CAST(SUM(R) AS FLOAT) / SUM(AB)  AS runs_scored_per_at_bat,
            CAST(SUM(SO) AS FLOAT) / SUM(AB) AS team_strikeouts_per_at_bat,
            SUM(b.'2B') + SUM(b.'3B') + SUM(HR) / CAST(SUM(AB) AS FLOAT) AS team_xbh_per_at_bat
          FROM batting AS b
          WHERE yearID > 1960
          GROUP BY yearID, teamID
          "

head(dbGetQuery(con, query))

dbGetQuery(con, query) %>%
  ggplot(aes(x = team_strikeouts_per_at_bat, y = runs_scored_per_at_bat)) +
  geom_point(aes(color = yearID)) +
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")

dbGetQuery(con, query) %>%
  ggplot(aes(x = team_xbh_per_at_bat, y = runs_scored_per_at_bat)) +
  geom_point(aes(color = yearID)) +
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")
```

SO I think this is interesting. It seems as though strikeouts per at bat has nearly no correlation with the amount of runs a team scores, while extra base hits has a decent, clear correlation.  

Let's see if this relationship holds in, say, the 2000s

```{r}
query <-  "
          SELECT 
            yearID,
            teamID,
            CAST(SUM(R) AS FLOAT) / SUM(AB)  AS runs_scored_per_at_bat,
            CAST(SUM(SO) AS FLOAT) / SUM(AB) AS team_strikeouts_per_at_bat,
            SUM(b.'2B') + SUM(b.'3B') + SUM(HR) / CAST(SUM(AB) AS FLOAT) AS team_xbh_per_at_bat
          FROM batting AS b
          WHERE yearID > 1999
          GROUP BY yearID, teamID
          "

head(dbGetQuery(con, query))

dbGetQuery(con, query) %>%
  ggplot(aes(x = team_strikeouts_per_at_bat, y = runs_scored_per_at_bat)) +
  geom_point(aes(color = yearID)) +
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")

dbGetQuery(con, query) %>%
  ggplot(aes(x = team_xbh_per_at_bat, y = runs_scored_per_at_bat)) +
  geom_point(aes(color = yearID)) +
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")
```

The correlation of strikeouts to runs is still weaker than the correlation between extra base hits and runs. Which points to the idea that teams/players would be willing to risk getting x more strikeouts if it came with getting x more extra base hits in a season.

```{r}
y <- dbGetQuery(con, query)$runs_scored_per_at_bat
x <- dbGetQuery(con, query)$team_strikeouts_per_at_bat

#Prints the R-squared of the regression of runs scored on strikeouts
summary(lm(y ~ x))$r.squared
```

```{r}
y <- dbGetQuery(con, query)$runs_scored_per_at_bat
x <- dbGetQuery(con, query)$team_xbh_per_at_bat

#Prints the R-squared of the regression of runs scored on extra base hits
summary(lm(y ~ x))$r.squared
```

There seems to be some legs here. Another trend I would expect to see is that batters who have increased the average exit velocities of their hits have gotten more extra base hits and also strike out more. And I would also expect to see the average exit velocity for hitters in the MLB to be increasing over time. So I need to see if I can find some new data to investigate these hypotheses.  

I am pretty sure the MLB started keeping stats like exit velocity around 2000, so I am going to do a little hunting on the internet.  

I found some data from [Fangraphs](https://www.fangraphs.com/leaders.aspx?pos=all&stats=bat&lg=all&qual=y&type=c,11,12,5,9,10,16,211&season=2019&month=0&season1=2008&ind=1&team=0&rost=0&age=0&filter=&players=0&startdate=2008-01-01&enddate=2019-12-31) from 2008-2019 which includes a Hard Hit %, which I think will act similarly to exit velocity in analysis.

```{r}
fangraphs_batting <- read_csv("fangraphs_batting.csv")

head(fangraphs_batting)

fangraphs_batting %>%
  group_by(Season) %>%
  summarise(count = n())
```

There are a similar number of observations for every season. Let's see if strikeouts/hard hit % have been increasing for this sample of players.

```{r}
advanced_batting <- fangraphs_batting %>%
  group_by(Season) %>%
  summarise(count = n(),
            so_per_ab = sum(SO) / sum(AB),
            avg_hard_percentage = mean(Hard_Percent))

advanced_batting %>%
  ggplot(aes(x = Season)) +
  geom_line(aes(y = so_per_ab, color = "strikeouts per at bat")) +
  geom_line(aes(y = avg_hard_percentage, color = "average hard hit ball per ball in play")) +
  ylab("percentage")
```
<br>
The strikeout pattern is not the same as the one for all players, but it is quite similar. Also, the hard hit percentage has risen dramatically, especially in the last few years, as the strikeout percentage has similarly been rising dramatically.  

```{r}
advanced_batting %>%
  ggplot(aes(x = avg_hard_percentage, y = so_per_ab)) +
  geom_point() +
  stat_smooth(geom='line', method = "lm", alpha=0.3, se=FALSE, color = "blue")

y <- advanced_batting$so_per_ab
x <- advanced_batting$avg_hard_percentage

#Prints the R-squared of the regression of strikeouts on hard hit balls
summary(lm(y ~ x))$r.squared
```
<br>
Can we see this pattern with individuals?  
This is going to take a bit of data manipulation.

```{r}
select_years_list <- fangraphs_batting %>%
  group_by(Name) %>%
  summarise(count = n()) %>%
  filter(count %in% c(8, 9))

select_years_list
```

I am going to use this group of players, some who are reaching the tail end of their career, and some who are reaching the peak years of their career. I would expect the correlation between changes in hard hit % and changes in strikeouts ber at bat to be fairly strong for most of these players, but stronger for those reaching the tail end of their career, as the younger players are still gaining strength, and are potentially able to hit more hard balls without changing their swing/approach.

```{r}
select_years_list$Name

fangraphs_batting %>%
  filter(Name %in% select_years_list$Name) %>%
  mutate(so_per_ab = SO/AB) %>%
  select(Name, Season, so_per_ab, Hard_Percent)

fangraphs_batting %>%
  filter(Name %in% select_years_list$Name) %>%
  mutate(so_per_ab = SO/AB) %>%
  select(Name, Season, so_per_ab, Hard_Percent) %>%
  ggplot(aes(x = Season)) +
  geom_line(aes(y = so_per_ab, color = "strikeouts per at bat")) +
  geom_line(aes(y = Hard_Percent, color = "Hard Hit Ball %")) +
  facet_wrap(~ Name)
```
<br>
This visualization did not help much, but it look pretty cool.  

I still think my hypothesis has legs. Just to make it crystal clear, my hypothesis is as follows:  
<br>
**A FUNDAMENTAL CHANGE IN HITTING MINDSET IN THE MLB IS RESPONSIBLE FOR THE HUGE INCREASE IN STRIKEOUTS IN RECENT YEARS. SPECIFICALLY, PLAYERS ARE CONSCIOUSLY SWINGING HARDER, KNOWING THAT THEY MAY WHIFF MORE OFTEN, BUT ALSO KNOWING THAT THEY WILL GET MORE EXTRA BASE HITS. THIS CHANGE IN MINDSET IS MOTIVATED BY GOAL OF SCORING AS MANY RUNS AS POSSIBLE, BECAUSE, ALL ELSE EQUAL, MORE RUNS MEANS MORE WINS.**

<br>
I want to make a graph that shows team strikeouts, extra base hits, and runs scored since 2000. Let's see if I can get that.

```{r}
query <-  "
          SELECT 
            yearID,
            teamID,
            SUM(SO) AS team_strikeouts,
            SUM(R) AS team_runs,
            SUM(b.'2B') + SUM(b.'3B') + SUM(HR) AS team_xbh
          FROM batting AS b
          WHERE yearID > 1999
          GROUP BY yearID, teamID
          "

head(dbGetQuery(con, query))

dbGetQuery(con, query) %>%
  ggplot(aes(x = yearID)) +
  geom_line(aes(y = team_runs, color = "team runs")) +
  geom_line(aes(y = team_xbh, color = "team extra base hits")) +
  facet_wrap(~ teamID)
```

OK this plot is unreal. I am going to add the teams that changed cities together in order to make it more uniform and even easier to visualize.

```{r}
dbGetQuery(con, query) %>%
  mutate(teamID = ifelse(teamID == "ANA", "LAA",
                         ifelse(teamID == "MON", "WAS",
                                ifelse(teamID =="FLO", "MIA", teamID)))) %>%
  ggplot(aes(x = yearID)) +
  geom_line(aes(y = team_runs, color = "team runs")) +
  geom_line(aes(y = team_xbh, color = "team extra base hits")) +
  facet_wrap(~ teamID)
```

Okay awesome. Now I am going to look at the same graph, but see how correlated strikeouts are with runs scored. As we saw in an earlier scatterplot, there should be little to no correlation evident.

```{r}
dbGetQuery(con, query) %>%
  mutate(teamID = ifelse(teamID == "ANA", "LAA",
                         ifelse(teamID == "MON", "WAS",
                                ifelse(teamID =="FLO", "MIA", teamID)))) %>%
  ggplot(aes(x = yearID)) +
  geom_line(aes(y = team_runs, color = "team runs")) +
  geom_line(aes(y = team_strikeouts, color = "team strikeouts")) +
  facet_wrap(~ teamID)
```

Pretty much every teams' strikeouts are increasing season over season, but some teams are not seeing the tradeoff expected with an increase in xbh and runs. For that reason I suspect that there may be another variable adding to the increase in strikeouts - pitcher skill.  

To me, it would make sense that if pitchers are outpacing batters in improving their talent year over year, the result would be more strikeouts for everyone. Let's see if we can determine what makes a pitcher induce more strikeouts using our Lahman pitching data.

```{r}
query <-  "
          SELECT *
          FROM pitching
          LIMIT 5
          "

dbGetQuery(con, query)
```
<br>
IPOuts appears to be a stat showing the number of outs that pitcher got that season. Therefore, SO/IPOuts would be the percentage of outs that were strikeouts. So I can run a fairly simple query that will give me the top strikeout pitchers of all time.

```{r}
query <-  "
          SELECT 
            pe.nameFirst || ' ' || pe.nameLast AS name,
            pe.debut AS first_game_date,
            SUM(GS) AS career_starts,
            SUM(IPouts) AS career_outs,
            SUM(SO) AS career_strikeouts,
            ROUND(CAST(SUM(SO) AS FLOAT) / SUM(IPouts), 3) AS strikeout_percentage
          FROM pitching AS pi
          INNER JOIN people AS pe ON pe.playerID = pi.playerID
          WHERE IPouts > 100
          GROUP BY pi.playerID
          ORDER BY strikeout_percentage DESC
          LIMIT 30
          "

kable_query(con, query)
```

Some clear observations here:

* These guys are pretty much all relief pitchers

* The list is heavily populated with current MLB pitchers

* Of the names I recognize, these pitchers throw HEAT

<br>

I have downloaded some pitching data [from Fangraphs](https://www.fangraphs.com/leaders.aspx?pos=all&stats=pit&lg=all&qual=y&type=c,7,8,13,-1,36,75,76&season=2019&month=0&season1=2007&ind=1&team=&rost=&age=&filter=&players=&startdate=&enddate=). Now I can look at the correlation between fastball speed and strikeouts more precisely.

```{r}
fangraphs_pitching <- read_csv("fangraphs_pitching")

head(fangraphs_pitching)
```

