---
title: "Deriving Insight From Lahman's Baseball Database"
author: "Michael Calabro"
date: "7/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<br>
Welcome to my journey_to_insight.Rmd file! In this markdown, I will be going step by step through my process of using the Lahman Baseball Database to seek answers to my question regarding Strikeouts in Major League Baseball.  
<br>  
<center>
**What is the Lahman Baseball Database? And what is your question about MLB strikeouts?**
</center>
<br>
The Lahman Baseball Database contains MLB statistics and data from 1871-2019...it contains nearly 250 years worth of MLB data to explore!  
I downloaded the SQLite database into my RSTUDIO Project, and can access all of its information with the following commands.

```{r, warning=FALSE, message=FALSE}
# RSQLite library allows me to use SQLite
# DBI library allows me to connect to the database
# tidyverse contains many libraries, most notably ggplot2 for visualizations
# kableExtra allows me to make my tables more presentable
library(RSQLite)
library(DBI)
library(tidyverse)
library(kableExtra)

# Now I need to establish my connection to the Lahman database
con <- dbConnect(SQLite(), 
                 dbname = "lahmans_baseball_db.sqlite")
```
<br>  
This "con" keyword now allows me to connect to the database. To view the list of tables in the database, I simply run the code below.  

```{r}
dbListTables(con)
```
<br>  
To view data from a specific table, I use the dbGetQuery function. The "batting" table will be key to my analysis, let's check it out.

```{r}
# First I write out the query and assign it to "query"
query <-  "
          SELECT *
          FROM batting
          LIMIT 5
          "

# Then I send it through the DBGetQuery function, along with the "con" connection, to run the query
dbGetQuery(con, query)

# And for tables that aren't so wide, the kable funtion with kable_styling makes the table very presentable
dbGetQuery(con, query) %>%
  kable() %>%
  kable_styling(full_width = FALSE, bootstrap_options = "bordered")

# I like the kable and think I'll use it a lot, so I am going to make a function kable_query for efficiency
kable_query <- function(con, query){
  return(
    dbGetQuery(con, query) %>%
      kable() %>%
      kable_styling(full_width = FALSE, bootstrap_options = "bordered")
  )
}
```
<br>  
As we can see, every row in the batter table consists of a player, a year, and all of the player's batting statistics in that year.  
<br>
<center>
**So What is your question about Strikeouts in the MLB?**
</center>
<br>  
A few years back (in 2018 I believe) I read an article which mentioned the fact that MLB players are striking out more and more every year, always breaking the single season record for total strikeouts. A more updated article, linked [here](https://www.espn.com/mlb/story/_/id/27694758/mlb-sets-strikeout-record-12th-straight-season), notes that this increase in strikeouts has been happening every year since 2008! Let's see if our Lahman data shows the same trend.

```{r}
query <-  "
          SELECT 
            yearID,
            SUM(SO) AS total_strikeouts
          FROM batting
          WHERE yearID > 2006
          GROUP BY yearID
          "

kable_query(con, query)

# Now I can use ggplot, with geom_point to make a scatterplot, geom_line to make a line graph, or both!
# Also, using %>% allows me to "pipe" my data into certain functions, like ggplot
dbGetQuery(con, query) %>%
  ggplot(aes(x = yearID, y = total_strikeouts)) +
  geom_point() +
  geom_line() +
  ggtitle("Yearly Total MLB Strikeouts From 2007-2019")
```

So what is causing this increase in strikeouts every year?? That is what I wish to investigate throughout the rest of this document and project.
