---
title: "Visualizing the Increase in MLB Strikeouts Since 2007"
author: "Michael Calabro"
date: "8/3/2020"
output: pdf_document
---

```{r setup, include=FALSE}
library(DBI)
library(tidyverse)
library(hexbin)

con <- dbConnect(RSQLite::SQLite(), dbname = "lahmans_baseball_db.sqlite")

# Makes it so all of my code chunks will appear in the pdf,
# and all of my code chunks written in sql will use the "con" connection
knitr::opts_chunk$set(connection = "con")
knitr::opts_chunk$set(echo = TRUE)
```

The goal of this document is to present, in a clean fashion, the most notable tables, graphs, and plots created in my "journey to insight". Graphs and tables which need explanations will be accompanied with explanations. And for now, I think we can dive right in!  

```{sql, tab.cap = "Total Strikeouts in the MLB Per Season", max.print = NA}
SELECT 
  yearID AS Year,
  SUM(SO) AS Strikeouts
FROM batting
WHERE yearID > 2006
GROUP BY yearID

-- Assigned to variable "total_so"
```

This table provides a clean representation of what I wish to investigate throughout this document.

```{sql, echo=FALSE, output.var="total_so"}
SELECT 
  yearID AS Year,
  SUM(SO) AS Strikeouts
FROM batting
WHERE yearID > 1994
GROUP BY yearID
```

\newpage

```{r}
total_so %>%
  ggplot(aes(x = Year, y = Strikeouts)) +
  geom_line() +
  xlim(1995, 2020) +
  ggtitle("Total Strikeouts in Each MLB Season Since 1995", 
          subtitle = "A new record has been set every year since 2007")
```

Of course, it is possible that this trend is due to an increase in at bats in this time frame. What I really want to know is if a higher *percentage* of at bats are resulting in strikeouts year over year.

\newpage

```{sql, tab.cap = "Strikeouts Percentage in the MLB By Season", max.print = NA}
SELECT 
  yearID AS Year,
  ROUND(CAST(SUM(SO) AS FLOAT) / SUM(AB), 3) AS "Strikeouts Per AB"
FROM batting
WHERE yearID > 2006
GROUP BY yearID

-- Assigned to variable strikeout_percent
```

```{sql, echo=FALSE, output.var="strikeout_percent"}
SELECT 
  yearID AS Year,
  ROUND(CAST(SUM(SO) AS FLOAT) / SUM(AB), 3) AS strikeouts_per_ab
FROM batting
WHERE yearID > 1994
GROUP BY yearID
```

\newpage

```{r}
strikeout_percent %>%
  ggplot(aes(x = Year, y = strikeouts_per_ab)) +
  geom_line() +
  xlim(1995, 2020) +
  ylab("Share of At Bats Resulting in Strikeouts") +
  ggtitle("Strikeouts Per At Bat Each MLB Season Since 1995", 
          subtitle = "A new record has been set every year since 2007")  
```

I find it very notable that this trend began around 2005 - shortly after the release of "Moneyball", a book which emphasized a more data-driven approach to winning baseball games. Since the release of "Moneyball", by Michael Lewis, strategy in the all aspects of the MLB has slowly begun to evolve. For this reason, it is my belief that this increase in strikeouts is likely due to an evolution in both hitting and pitching strategy for MLB teams. To make it more specific, I have two theories:  

1. Batters/ team batting strategies have changed to incentivize an increase in some metric(s) (i.e walks, homeruns, extra base hits, exit velocity) at the expense of more strikeouts, for the sake of more runs.  

*Simply: Batters changing strategy = more strikeouts*

2. Pitchers/ team pitching strategies have outpaced hitting strategies in their ability to learn from the new advanced metrics.  

*Simply: Pitchers changing strategy = more strikeouts*  

So to what extent is each theory accurate? To find out, I need to come up with some investigatable hypotheses, and then fudge around with the data available to judge their respective merit.

\newpage

**BATTING HYPOTHESIS #1: Some Hitting Statistic is Increasing Alongside Strikeouts**

```{sql, tab.cap = "Hitting Percentages in the MLB By Season", max.print = NA}
SELECT 
  yearID AS Year,
  ROUND(CAST(SUM(SO) AS FLOAT) / SUM(AB), 3) AS "SO%",
  ROUND(CAST(SUM(H) AS FLOAT) / SUM(AB), 3) AS "Hit%",
  ROUND(CAST(SUM(HR) AS FLOAT) / SUM(AB), 3) AS "HR%",
  ROUND(CAST(SUM('2B') AS FLOAT) / SUM(AB), 3) AS "2B%",
  ROUND(CAST(SUM('3B') AS FLOAT) / SUM(AB), 3) AS "3B%",
  ROUND(CAST(SUM('3B') + SUM('2B') + SUM(HR) AS FLOAT) / SUM(AB), 3) AS "XBH%",
  ROUND(CAST(SUM(BB) AS FLOAT) / SUM(AB), 3) AS "Walk%"
FROM batting
WHERE yearID > 2006
GROUP BY yearID
```

While an emphasis is often placed on Homeruns, it seems to me like the most notable trend is an increase in the percentage of all extra base hits (2B, 3B, and HR are all XBHs).  

Furthermore, I want to see extra base hits as a share of all hits. The logic goes that players are willing to strike out more, if it means that more of their hits are extra base hits.

\newpage

```{sql, tab.cap = "Select Hitting Percentages in the MLB By Season", max.print = NA}
SELECT
  yearID AS Year,
  ROUND(CAST(SUM(SO) AS FLOAT) / SUM(AB), 3) AS "SO per AB",
  ROUND((SUM('3B') + SUM('2B') + SUM(HR)) / CAST(SUM(H) AS FLOAT), 3) AS "XBH per Hit"
FROM batting
WHERE yearID > 2006
GROUP BY yearID

-- Assigned to variable so_xbh
```

```{sql, echo=FALSE, output.var="so_xbh"}
SELECT
  yearID AS Year,
  ROUND(CAST(SUM(SO) AS FLOAT) / SUM(AB), 3) AS so_per_ab,
  ROUND((SUM('3B') + SUM('2B') + SUM(HR)) / CAST(SUM(H) AS FLOAT), 3) AS xbh_per_hit
FROM batting
WHERE yearID > 2006
GROUP BY yearID
```

\newpage

```{r}
so_xbh %>%
  ggplot(aes(x = Year)) +
  geom_line(aes(y = so_per_ab, color = "Strikeouts Per At Bat")) +
  geom_line(aes(y = xbh_per_hit, color = "Extra Base Hits as a Share of Hits")) +
  xlab("percentage") +
  ggtitle("Extra Base Hits and Strikeouts in the MLB Since 2007")
```

I plan to do more statistical analysis in another document, but I think I see a pretty clear pattern here.  

The next question is: why would someone want to sacrifice strikeouts for extra base hits?  

Answer: The offensive objective in baseball is to score as many runs as possible. Therefore, I would guess that team extra base hits play a bigger role in runs scored than strikeouts do. If that were the case, then teams would be incentivized to risk strikeouts for the sake of extra base hits, becasue then they would be likely to score more runs.

\newpage

**BATTING HYPOTHESIS #2: XBH Has a Greater Effect on Runs Scored Than Strikeouts**

```{sql, output.var="team_xbh_so_runs"}
SELECT 
  yearID,
  teamID,
  SUM(SO) AS team_strikeouts,
  SUM(R) AS team_runs,
  SUM(b.'2B') + SUM(b.'3B') + SUM(HR) AS team_xbh
FROM batting AS b
WHERE yearID > 1997
GROUP BY yearID, teamID

-- Assigned to variable "team_xbh_so_runs"
```

```{r}
team_xbh_so_runs %>%
  # Mutate combines teams that moved cities in this time frame into the same plot
  mutate(teamID = ifelse(teamID == "ANA", "LAA",
                         ifelse(teamID == "MON", "WAS",
                                ifelse(teamID =="FLO", "MIA", teamID)))) %>%
  ggplot(aes(x = yearID)) +
  geom_line(aes(y = team_runs, color = "team runs")) +
  geom_line(aes(y = team_xbh, color = "team extra base hits")) +
  facet_wrap(~ teamID) +
  theme(axis.text.x=element_blank()) +
  xlab("Year (1998-2019)") +
  ylab("count") +
  ggtitle("MLB Team Extra Base Hits and Runs From 1998-2019")
```

\newpage

```{r}
team_xbh_so_runs %>%
  # Mutate combines teams that moved cities in this time frame into the same plot
  mutate(teamID = ifelse(teamID == "ANA", "LAA",
                         ifelse(teamID == "MON", "WAS",
                                ifelse(teamID =="FLO", "MIA", teamID)))) %>%
  ggplot(aes(x = yearID)) +
  geom_line(aes(y = team_runs, color = "team runs")) +
  geom_line(aes(y = team_strikeouts, color = "team strikeouts")) +
  facet_wrap(~ teamID) +
  theme(axis.text.x=element_blank()) +
  xlab("Year (1998-2019)") +
  ylab("count") +
  ggtitle("MLB Team Strikeouts and Runs From 1998-2019")
```

To my eye, it seems fairly clear that runs follow trend with extra base hits much more closely than they do with strikeouts. For that reason, it seems to make sense that coaches would want their hitters attempting to hit more extra base hits, even if that means striking out more often.  

That being said, it is also fairly clear that nearly every team is striking out MUCH more in recent years, even if they aren't seein the accompanying rise in xbh and runs. This tells me that improved pitching across the MLB may also be playing a role in these crazy strikeout numbers.

\newpage

```{r}
runs = team_xbh_so_runs$team_runs
xbh = team_xbh_so_runs$team_xbh
so = team_xbh_so_runs$team_strikeouts

summary(lm(runs ~ xbh + so))$coefficients %>%
  kableExtra::kable() %>%
  kableExtra::kable_styling(full_width = FALSE)
summary(lm(runs ~ xbh + so))$r.squared
```

As you can see from the regression, an increase in 1 extra base hit leads to an expected 1.4 runs added in a season, while an increase in 1 strikeout leads to an expected .137 runs lost. With this in mind, it would make sense that, if a team/player could get some more extra base hits at the expense of striking out more, they would choose to do so.
